Sequence Specific Modeling of Cell-Free Protein Synthesis

### 2017-02-28 Paper_v1.tex
supportinfo

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\author{Nicholas Horvath}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1~(607)~255-4258}
\fax{+1~(607)~255-9166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields; in addition, the translation rate is the next step to optimize for higher productivity and yield.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
\end{abstract}

@@

Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E. coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems using linear programming.
Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}. Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
\begin{figure}[t!]
\centering
\includegraphics[width=0.6\textwidth]{./Figures/GFP_plasmid.pdf}
\caption{Predicted deGFP concentration (black line) at different plasmid concentrations versus measurements of deGFP (dots) synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:GFP_plasmid}
\end{figure}

These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
In all cases, CFPS was supplied with glucose.
In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (control).
In the second case, CFPS was supplied with amino acids, but the amino acid synthesis was turned off (AA uptake w/o synthesis).
We shut off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).
Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system had sufficient substrates to power the system and synthesize each protein.
The third case had the lowest productivity for each protein; thus, the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.  
The qualitative trend of productivity for the three different cases was the same within each case; however, some proteins had higher productivity than others.
For instance, in the second case, FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system relied on a mixture of glucose and some amino acids for the production of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 22.2\%.
Only the necessary amount of amino acids were used for the production of deGFP; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case it is 92.5\%.
Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 14.8\% and the system used about twice the amount of glucose as in the first two cases.
In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
Interestingly, CAT carbon yield showed similar performance for all three cases; however, it still had a lower yield than the first two cases.
Each protein had different energy requirements for transcription and translation depending on its sequence.
In the case of CAT, energy requirements for transcription or translation were low enough that its carbon yield was not hindered in the third case as significantly as for the other proteins.

\begin{table}[t!]
\centering
    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
    \renewcommand{\arraystretch}{0.9}
    \begin{tabular}{lccc} \toprule
        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Produced (mM/hr)} & Control & \parbox{2.6 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Consumed (mM/hr)} & \\ \midrule
        GLC & 37.3 & 33.7 & 66.9 \\
        ALA & 0.0 & 0.2 & - \\
        ARG & 0.3 & 0.3 & - \\
        ASN & 0.5 & 0.4 & - \\
        ASP & 0.4 & 0.6 & - \\
        CYS & 0.2 & 0.1 & - \\
        GLU & 2.1 & 0.6 & - \\
        GLN & 0.4 & 0.3 & - \\
        GLY & 0.3 & 0.3 & - \\
        HIS & 0.5 & 0.5 & - \\
        ILE & 0.0 & 0.6 & - \\
        LEU & 1.0 & 1.0 & - \\
        LYS & 1.0 & 0.9 & - \\
		MET & 0.0 & 0.2 & - \\
        PHE & 1.0 & 0.9 & - \\
        PRO & 0.4 & 0.4 & - \\
        SER & 0.0 & 0.2 & - \\
        THR & 1.0 & 0.5 & - \\
        TRP & 0.1 & 0.1 & - \\
        TYR & 0.8 & 0.8 & - \\
        VAL & 0.6 & 0.7 & - \\ \hline
        Sum & 47.9 & 43.3 & 66.9 \\ \toprule
        Yield & 20.5\% & 22.2\% & 14.8\% \\
	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
    \end{tabular}
\label{tbl:yield_breakdown}
\end{table}

We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield_POI}B).
The same inverse qualitative trend is observed as for productivity, but it was less significant.
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
As the protein length increased, the productivity and carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
We thus wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS performance.
%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
%In addition, ssFBA is solved by setting the production of the protein as the objective function.
%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Carbon yield (Error bars represent the 95\% CI of the ensemble). B. Mean carbon yield versus the carbon number for each corresponding protein.}
\label{fig:Yield_POI}
\end{figure}

\subsection{CFPS performance sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the elongation rate by ribosomes being the most significant.
This showed that translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription rates.
This is consistent with literature findings.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, when amino acid synthesis was removed, amino acid uptake became more important, as it was the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray). (Error bars represent 95\% CI of the ensemble).}
\label{fig:SI_GFP}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
\author{Nicholas Horvath}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
\author{Joseph Wayman}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
\author{Kara Calhoun}
\affiliation[Stanford University]
{School of Chemical Engineering, Stanford University, Stanford, CA 94305}
\author{James Swartz}
\affiliation[Stanford University]
{School of Chemical Engineering, Stanford University, Stanford, CA 94305}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1~(607)~255-4258}
\fax{+1~(607)~255-9166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, cell-free protein synthesis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity, energy efficiency, and carbon yield.
A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate is the next step to optimize for higher productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
\end{abstract}

@@

Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E. coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}. Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 272 reactions and 145 species (Fig \ref{fig:network}A).
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:network}B).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:network}C).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:network}D).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_network.pdf}
\caption{Sequence specific flux balance analysis. A. Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway. Thick gray arrows indicate withdrawal of precursors for amino acid synthesis. B. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. D. Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
\label{fig:network}
\end{figure}

These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with few adjustable parameters in the promoter model taken from literature \cite{Moon:2012aa}.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.

Our next goal was to examine the performance of CFPS for eight different proteins under three different cases. 
Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
In all cases, CFPS was supplied with glucose.
In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (AA uptake and synthesis).
In the second case, CFPS was supplied with amino acids, but the amino acid synthesis were turned off (AA uptake w/o synthesis).
These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).  
Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod}).
An additional case was considered for CAT, since a comprehensive dataset is available \cite{2005_calhoun_BiotechnologyProgress}; thus fluxes were constrained to experimental measurements where avaliable, except for CAT production which was determined by the transcription/translation parameters.

\subsection{CFPS Productivity}
We evaluated CFPS productivity for all eight proteins and for each case. 
All cases had very similar performance for each protein and were in the standard deviation of each other (Fig \ref{fig:Prod}A). 
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.  
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 ($\mu$M/h) whereas CAT had a productivity of about 12 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information). 
CAT was an outlier for the trendline, even though it was in the same order of magnitude of the trendline.
The difference in the T7 and  $\sigma_{70}$ promoters did not alter the qualitative productivity performance of CAT. 
The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation, however this should be investigated further.
% a further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated. 
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Productivity.pdf}
\caption{CFPS productivity of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white). A. Productivity (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein. Single trendline of productivity versus carbon number (dotted line) for all cases (R\textsuperscript{2} = 0.99). Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Prod}
\end{figure}

The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.

\subsection{CFPS Energy Efficiency}
Following the same outline as in examining the productivity, we calculated the energy efficiency for each protein (Fig \ref{fig:Energy}A).
The first two cases had comparable performance with having the highest energy efficiency, where amino acids were supplied in the media.
The third case (with no amino acid uptake) had the lowest energy efficiency; this is because glucose is utilized to synthesize the necessary amino acids for each protein as well as energize the system. 
We next investigated the effect of the carbon number of each protein on the energy efficiency (Fig \ref{fig:Energy}B).
The same inverse qualitative trend is observed as for productivity, but followed a linear trendline.
The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for the first two cases.
Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, thus it makes sense for larger proteins to have a lower energy efficiency.  
The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information). 
Interestingly, in the third case each protein had a similar energy efficiency of about 25\% regardless of the carbon number of the protien. 
In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein of interest kept the energy efficiency saturated at a relatively low level.
However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.9 $\pm$ 1.5\% compared to the theoretical maximum of 72 $\pm$ 7.1\%. 
This shows CFPS systems have a lot of room for improvement: first, the experimental setup was still producing certain amino acids which could be turned off. 
Second, the system has a high accumulation of metabolic byproducts of organic acids which is a result of inefficient energy utilization.

%This may be a potential area for improvement in CFPS to optimize for energy utilization. 
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
\caption{CFPS energy efficiency of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white). A. Energy efficiency (Error bars represent the 95\% CI of the ensemble). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.97) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.79). Dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Energy}
\end{figure}

\subsection{CFPS Carbon Yield}
We also calculated the carbon yield for each protein (Fig \ref{fig:Yield}A).
The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 56.9 $\pm$ 2.4\% for CAT.
Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 57.0 $\pm$ 2.2\% for CAT.
Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution increases to 100\%.
Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\%  for CAT and the system used about twice the amount of glucose as in the first two cases.
In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins.
The experimentally constrained case for CAT had carbon yield of about 6\% compared to the theoretical maximum of 57\%.
This loss in carbon yield shows inefficiencies in CFPS that can be potentially improved.
Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
In addition, ssFBA is solved by setting the production of the protein as the objective function.
Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate.
The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.    
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CarbonYield.pdf}
\caption{CFPS carbon yield of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white). A. Energy efficiency (Error bars represent the 95\% CI of the ensemble). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.92) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.83). Dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Yield}
\end{figure}
We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield}B).
The same inverse qualitative trend is observed as for productivity.
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
A single trendline was formulated for the first two cases since they had similar performance and another trendline was formulated for the third cases (available in Supporting Information)
As the protein size increased, the carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
